---
title: "SEC2_APM1110-FA1 GROUP_8-DABU, L; LIBRES, B"
subtitle: "Github: https://github.com/dabulira20/Prob-ProbDi-FAs/blob/main/FA1.Rmd"
output: pdf_document
date: "`r Sys.Date()`"
--- 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Problem 1
```
Write the skewness program, and use it to calculate the skewness coefficient of the four examination subjects in results.txt (results.csv). What can you say about these data? Is Pearson’s approximation reasonable?
```

```{r}
results <- read.csv("results.csv", na.strings = "NA")

# exact skewness (Equation 2.1)
exact_skewness <- function(x) {
  x <- x[!is.na(x)]
  n <- length(x)
  m <- mean(x)
  m2 <- sum((x - m)^2) / n
  m3 <- sum((x - m)^3) / n
  m3 / sqrt(m2)^3
}

# Pearson's approximate skewness: 3*(mean - median) / sd
pearson_skewness <- function(x) {
  x <- x[!is.na(x)]
  3 * (mean(x) - median(x)) / sd(x)
}

subjects <- c("arch1", "prog1", "arch2", "prog2")

skewness_results <- data.frame(
  arch1 = c(exact_skewness(results$arch1), pearson_skewness(results$arch1)),
  prog1 = c(exact_skewness(results$prog1), pearson_skewness(results$prog1)),
  arch2 = c(exact_skewness(results$arch2), pearson_skewness(results$arch2)),
  prog2 = c(exact_skewness(results$prog2), pearson_skewness(results$prog2))
)
rownames(skewness_results) <- c("Exact Skewness", "Pearson Skewness")
print(skewness_results)
```

#### What can you say about these data?

`arch1`, `prog1`, and `prog2` are negatively (left) skewed, meaning
most students scored fairly well with a smaller tail of low scorers.
`arch2` is positively (right) skewed instead, meaning most students
clustered at lower scores with a few high scorers pulling the tail up.
None of the subjects is heavily skewed — all four values are well
under 1 in magnitude.

#### Is it a reasonable approximation?

Pearson's formula gets the sign right for all four subjects, correctly
flagging `arch2` as the only right-skewed one. The magnitudes are
somewhat close but generally larger than the exact values (e.g.
`prog1`: -0.63 vs -0.34), so it's a reasonable quick approximation but
not a substitute for the exact formula when precision matters.


### Problem 2
```
For the class of 50 students of computing detailed in Exercise 1.1, use R to
(a)  form the stem-and-leaf display for each gender, and discuss the advantages of this representation compared to the traditional histogram; 
(b)  construct a box-plot for each gender and discuss the findings. 
```
#### Answers:
##### (a) Stem-and-Leaf Display by Gender

```{r}
female <- c(57, 59, 78, 79, 60, 65, 68, 71, 75, 48, 51, 55, 56, 41, 43,
            44, 75, 78, 80, 81, 83, 83, 85)
male <- c(48, 49, 49, 30, 30, 31, 32, 35, 37, 41, 86, 42, 51, 53, 56, 42, 44, 
          50, 51, 65, 67, 51, 56, 58, 64, 64, 75)

print("First-Year Java Programming Exam Scores of Female Students")
stem(female)
print("First-Year Java Programming Exam Scores of Male Students")
stem(male)

```

**Advantages of stem-and-leaf over histogram:**
A stem-and-leaf plot retains the actual data values while still showing the
shape of the distribution. Unlike a histogram, you can read individual scores
directly from the display, making it easier to spot outliers and check symmetry
without losing information.

##### (b) Box Plots by Gender

```{r}
boxplot(female, male,
        names = c("Females", "Males"),
        main = "First-Year Java Programming Exam Scores",
        ylab = "Scores",
        ylim = range(0, 100),
        col = c("pink", "lightblue"))

```

#### Discussion of Findings: 
- The box for females sits higher on the chart than the box for males, so females generally scored better on the exam.

- The female box looks taller than the male box, meaning female scores were more spread out, while male scores were more bunched or compressed together.

- There's a dot above the male box (an outlier), showing one male student scored much higher than the rest of the male group. The female box has no dots like this, meaning there were no extreme outliers in the female scores.

- Overall, the boxplot shows females scored higher on average but with more variation, while males scored lower but more consistently.
