---
title: "SEC2_APM1110-FA2-DABU, L."
subtitle: "GitHub: https://github.com/dabulira20/Prob-ProbDi-FAs/tree/main/FA2"
date: "`r Sys.Date()`"
output: 
  pdf_document: 
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### 1. Problem (#5) 
A geospatial analysis system has four sensors supplying images. The
percentage of images supplied by each sensor and the percentage of relevant
images (relative to a query) from each sensor are given below.

| Sensor | % of Images Supplied | % of Relevant Images |
|:------:|:---------------------:|:---------------------:|
|   1    |          15           |          50           |
|   2    |          20           |          60           |
|   3    |          25           |          80           |
|   4    |          40           |          85           |

We are asked for the **overall percentage of relevant images**.

#### Solution

Let $B_i$ be the event that an image comes from sensor $i$, and let $A$ be the
event that an image is relevant. We are given

$$P(B_1)=0.15,\; P(B_2)=0.20,\; P(B_3)=0.25,\; P(B_4)=0.40$$
$$P(A\mid B_1)=0.50,\; P(A\mid B_2)=0.60,\; P(A\mid B_3)=0.80,\; P(A\mid B_4)=0.85$$

By the **Law of Total Probability**:

$$P(A)=\sum_{i=1}^{4} P(B_i)\,P(A\mid B_i)$$

```{r item5-calc}
# Percentage of images supplied by each sensor (as proportions)
P_B <- c(0.15, 0.20, 0.25, 0.40)

# Percentage of relevant images from each sensor (as proportions)
P_A_given_B <- c(0.50, 0.60, 0.80, 0.85)

# Individual contribution of each sensor to overall relevance
contribution <- P_B * P_A_given_B
names(contribution) <- paste("Sensor", 1:4)
contribution

# Overall probability (percentage) of relevant images
P_A <- sum(contribution)
P_A
```

```{r item5-percent}
cat("Overall percentage of relevant images =", P_A * 100, "%\n")
```

**Answer.**

$$P(A) = (0.15)(0.50)+(0.20)(0.60)+(0.25)(0.80)+(0.40)(0.85)$$
$$P(A) = 0.075+0.12+0.20+0.34 = 0.735$$

The overall percentage of relevant images is **73.5%**.

---

### 2. Problem (#6) 

A fair coin is tossed twice. Let

- $E_1$ = event that both tosses have the same outcome, $E_1=\{HH,TT\}$
- $E_2$ = event that the first toss is a head, $E_2=\{HH,HT\}$
- $E_3$ = event that the second toss is a head, $E_3=\{TH,HH\}$

Show that $E_1,E_2,E_3$ are **pairwise independent** but **not mutually
independent**.

#### Solution

The sample space for two coin tosses is
$$S=\{HH,\,HT,\,TH,\,TT\}$$
with each outcome equally likely, i.e. probability $1/4$ each.

```{r item6-setup}
# Sample space
S <- c("HH", "HT", "TH", "TT")
p_outcome <- 1/4  # each outcome equally likely

# Define the events
E1 <- c("HH", "TT")   # both tosses same
E2 <- c("HH", "HT")   # first toss is head
E3 <- c("TH", "HH")   # second toss is head

P_E1 <- length(E1) * p_outcome
P_E2 <- length(E2) * p_outcome
P_E3 <- length(E3) * p_outcome

c(P_E1 = P_E1, P_E2 = P_E2, P_E3 = P_E3)
```

Each event has probability $1/2$:
$$P(E_1)=P(E_2)=P(E_3)=\frac{2}{4}=\frac{1}{2}$$

### Pairwise independence

Two events $X$ and $Y$ are independent if $P(X\cap Y)=P(X)P(Y)$.

```{r item6-pairwise}
# Intersections
E1_E2 <- intersect(E1, E2)
E1_E3 <- intersect(E1, E3)
E2_E3 <- intersect(E2, E3)

P_E1E2 <- length(E1_E2) * p_outcome
P_E1E3 <- length(E1_E3) * p_outcome
P_E2E3 <- length(E2_E3) * p_outcome

data.frame(
  Pair = c("E1 & E2", "E1 & E3", "E2 & E3"),
  Intersection = c(paste(E1_E2, collapse=","),
                    paste(E1_E3, collapse=","),
                    paste(E2_E3, collapse=",")),
  P_intersection = c(P_E1E2, P_E1E3, P_E2E3),
  Product_of_P = c(P_E1*P_E2, P_E1*P_E3, P_E2*P_E3)
)
```

- $E_1\cap E_2=\{HH\}\Rightarrow P(E_1\cap E_2)=\dfrac{1}{4}=P(E_1)P(E_2)=\dfrac12\cdot\dfrac12=\dfrac14$ ✓
- $E_1\cap E_3=\{HH\}\Rightarrow P(E_1\cap E_3)=\dfrac{1}{4}=P(E_1)P(E_3)=\dfrac14$ ✓
- $E_2\cap E_3=\{HH\}\Rightarrow P(E_2\cap E_3)=\dfrac{1}{4}=P(E_2)P(E_3)=\dfrac14$ ✓

Since all three pairwise products match the corresponding joint
probabilities, $E_1,E_2,E_3$ are **pairwise independent**.

### Mutual independence check

For **mutual** independence, in addition to the pairwise condition we also
need
$$P(E_1\cap E_2\cap E_3)=P(E_1)\,P(E_2)\,P(E_3)$$

```{r item6-mutual}
E123 <- intersect(intersect(E1, E2), E3)
P_E123 <- length(E123) * p_outcome
product_all <- P_E1 * P_E2 * P_E3

data.frame(
  Quantity = c("P(E1 ∩ E2 ∩ E3)", "P(E1)P(E2)P(E3)"),
  Value = c(P_E123, product_all)
)
```

$$E_1\cap E_2\cap E_3=\{HH\}\Rightarrow P(E_1\cap E_2\cap E_3)=\frac14$$
$$P(E_1)P(E_2)P(E_3)=\frac12\cdot\frac12\cdot\frac12=\frac18$$

Since
$$\frac14 \neq \frac18,$$
the mutual-independence condition **fails**.

#### Conclusion

$E_1, E_2, E_3$ satisfy $P(E_i\cap E_j)=P(E_i)P(E_j)$ for every pair
$(i,j)$, so they are **pairwise independent**. However, because
$P(E_1\cap E_2\cap E_3)\neq P(E_1)P(E_2)P(E_3)$ ($\tfrac14 \neq \tfrac18$),
the three events are **not mutually independent**.
